NOPT048 Notes

### 2019-12-11 misc/study-notes/groups.tex
Added more isomorphism theorems

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}

@@

\section 1. PRELIMINARY NOTIONS FROM LINEAR ALGEBRA

This set of notes assumes that the reader has had some exposure to linear algebra; the most crucial notions will be outlined in this section.

@@

\section 2. GROUPS

\section 2.1. Properties and Basic Examples

@@

\section 2.2. Permutation Groups

@@

\section 2.3. Subgroups

@@

\section 2.4. Cosets and Normal Subgroups

@@

\section 2.5. Isomorphisms and Homomorphisms

\section 2.6. The Isomorphism Theorems

Now for any surjective map $f$ from sets $G$ to $H$, we can define an equivalence relation on elements $a,b\in G$:
$$a\sim b\quad \hbox{if and only if}\quad f(a) = f(b)$$
The set of equivalence classes is denoted $G/{\sim}$ and there exists a bijection $f':G/{\sim} \rightarrow H$. Now if $f$ is a group homomorphism, then we can pull back the group structure of $H$ to $G/{\sim}$ using the bijection. Then the set
$$G/K = \{gK : g\in G\}.$$
has a group structure and is called the {\it quotient group}. Multiplication in the quotient group is given by $gK\cdot hK = (g\cdot h) K$, which is well-defined because $K$ is normal. So the group structure of $G/K$ depends only on the kernel $K$ and not the homomorphism that induced it. The results of this section are summed up in the following theorem.

\parenproclaim Theorem I (First Isomorphism Theorem). Let $f:G\rightarrow H$ be a surjective group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug

The second and third isomorphism theorems give further results on the relationships between specific types of quotient groups.

\section 2.7. Inner and Outer Automorphisms

\proof Let $f_a\in Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
$$g\circ f_a\circ g^{-1} = g(a\cdot g^{-1}(s)\cdot a^{-1}) = g(a)\cdot s\cdot g(a)^{-1},$$
and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in Inn(G)$. So $\Inn(G)$ is stable under conjugation in $Aut(G)$.\slug

\section ?.?. Simplicity of the Alternating Group

A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$. In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple. First we establish that the set of 3-cycles generate $A_n$.

@@

\section ?.? Direct Products and Abelian Groups

@@

Let $G$ be a group and $S$ a set of elements. We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$. For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
$$O_s = \{g(s)\in S : g\in G\}$$
If there exists an $s$ such that $O_s = S$, then we say that $G$ acts {\it transitively} on $S$. In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $g(s) = s'$.

The {\it stabiliser} of $s$ is the set of all elements in $G$ that leave $s$ fixed:
$$G_s = \{g\in G : g(s) = s\}$$

Consider a model example of a transitive action. Let $G$ be a group with a subgroup $H\leq G$; let $S$ be the set $\{aH : a\in G\}$ of left cosets of $H$. The group $G$ acts on $S$ by taking a coset $aH$ to $gaH$. This action is transitive: For any $a, a'$, there exists some $g$ such that $g(a) = a'$, so we can take any $aH$ to $a'H$ via $g$ as well. The closure property of subgroups gives us $G_H = H$, and the stabiliser of an arbitrary coset $aH$ is the set $\{g\in G : gaH = aH\}$. We find that this is exactly the conjugate subgroup $aHa^{-1}$. More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subseteq G$.

@@

[Sylow's Theorems.]

>>>>

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Stab{\mathop{\rm Stab}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}

@@

\section 1. INTRODUCTION

This set of notes assumes that the reader has had some exposure to linear algebra; the most crucial notions are briefly outlined here. Next, the group axioms are introduced and some basic properties of groups are given.

@@

\section 1.3. Groups

@@

\section 1.4. Permutation Groups

@@

\section 1.5. Subgroups

@@

\section 2. NORMAL SUBGROUPS AND HOMOMORPHISMS

\section 2.1. Cosets and Normal Subgroups

@@

\section 2.2. Simplicity of the Alternating Group

We take a brief detour to prove a famous result in the classification of groups, using only the basic techniques we have learned so far. A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$. In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple. First we establish that the set of 3-cycles generate $A_n$.

@@

With this theorem in hand, it becomes easy to check that $A_4$ is the only alternating group that possesses a normal subgroup.

\section 2.3. Isomorphisms and Homomorphisms

\section 2.4. The Isomorphism Theorems

Now for any surjective map $f$ from sets $G$ to $H$, we can define an equivalence relation on elements $a,b\in G$:
$$a\sim b\quad \hbox{if and only if}\quad f(a) = f(b)$$
The set of equivalence classes is denoted $G/{\sim}$ and there exists a bijection $f':G/{\sim} \rightarrow H$. Now if $f$ is a group homomorphism, then we can pull back the group structure of $H$ to $G/{\sim}$ using the bijection. Then the set
$$G/K = \{gK : g\in G\}.$$
has a group structure and is called the {\it quotient group}. Multiplication in the quotient group is given by $gK\cdot hK = (g\cdot h) K$, which is well-defined because $K$ is normal. So the group structure of $G/K$ depends only on the kernel $K$ and not the homomorphism that induced it. The following theorem sums up what we have discovered so far.

\parenproclaim Theorem I (First Isomorphism Theorem). Let $f:G\rightarrow H$ be a surjective group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group isomorphism $f':G/K \rightarrow H$.\slug

This result is fundamental and is often used in conjunction with other laws of homomorphisms to establish isomorphisms. The other two isomorphism theorems provide further insight into the properties of group homomorphisms.

\parenproclaim Theorem J (Second Isomorphism Theorem). Let $G$ be a group with subgroups $A$ and $B$. Suppose that $A\leq N_G(B)$. Then $AB$ is a group and $B$ is a normal subgroup of $AB$. Then there exists a surjective homomorphism $f:A\rightarrow AB/B$ with kernel $A\cap B$, given by $f(a) = aB$.

\proof First we show that $AB$ is a group. Since $A\leq N_G(B)$, for any product $a_1b_1a_2b_2$, we can find $b'\in B$ such that $a_2b' = b_1a_2$. Then $a_1b_1a_2b_2 = a_1a_2b'b_2\in AB$. Similarly, $AB$ is stable under inverses, since for any product $ab$, we can find $b'\in B$ such that $(ab)^{-1} = b^{-1}a^{-1} = a^{-1}b'$. Because both $A$ and $B$ are subgroups of $N_G(B)$, so is $AB$ and in particular, $B$ is a normal subgroup of $AB$. Then $f:A\rightarrow AB/B$ is the restriction of the quotient map $AB\rightarrow AB/B$ to $A$, so it is surjective.

Now let $a\in \ker(f)$. By definition, $f(a) = aB = B$, so $a\in B$. Thus $a\in A\cap B$. Now let $a\in A\cap B$. Then $aB = B$ so $a\in \ker(f)$. Thus $\ker(f) = A\cap B$. In fact, we can now apply the first isomorphism theorem to deduce that $A/\ker(f) \cong AB/B$. \slug

\parenproclaim Theorem K (Third Isomorphism Theorem). Suppose that $A\leq B\leq G$ is a chain of subgroups and furthermore, both $A$ and $B$ are normal subgroups of $G$. Then the kernel of the surjective homomorphism $f:G/A\rightarrow G/B$ is $B/A$.

\proof The homomorphism $f$ is well defined, since if $sA = tA$, then $t^{-1}s\in A$, implying that $t^{-1}s\in B$ and $sB = tB$. It is also surjective, since for any $sB\in G/B$, we have $sA\in G/A$ and $f(sA) = sB$.

Now let $sA\in \ker(f)$. We have $f(sA) = sB = B$, so $s\in B$. This means that $sA\in B/A$. Conversely, if $s\in B$ then $f(sA) = sB = B$. So $\ker(f) = B/A$ and by the first isomorphism theorem, we can express this as the isomorphism $(G/A)/(B/A)\cong G/B$.\slug

[Include correspondence theorem.]

\section 3. AUTOMORPHISMS

\section 3.1. Inner and Outer Automorphisms

\proof Let $f_a\in \Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
$$g\circ f_a\circ g^{-1} = g(a\cdot g^{-1}(s)\cdot a^{-1}) = g(a)\cdot s\cdot g(a)^{-1},$$
and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in \Inn(G)$. So $\Inn(G)$ is stable under conjugation in $\Aut(G)$.\slug

\section ?.? Structure of Abelian Groups

@@

Let $G$ be a group and $S$ a set of elements. We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or $gs$) such that, for all $s\in S$, $es = s$ and $(gh)s = g\big(hs\big)$ for every $g,h\in G$. For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
$$Gs = \{gs\in S : g\in G\}$$
If there exists an $s$ such that $Gs = S$, then we say that $G$ acts {\it transitively} on $S$. In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $gs = s'$.

The {\it stabiliser} of $s$ is the set of all elements in $G$ that leave $s$ fixed:
$$\Stab(s) = \{g\in G : gs = s\}$$

Consider a model example of a transitive action. Let $G$ be a group with a subgroup $H\leq G$; let $S$ be the set $\{aH : a\in G\}$ of left cosets of $H$. The group $G$ acts on $S$ by taking a coset $aH$ to $gaH$. This action is transitive: For any $a, a'$, there exists some $g$ such that $g(a) = a'$, so we can take any $aH$ to $a'H$ via $g$ as well. The closure property of subgroups gives us $G_H = H$, and the stabiliser of an arbitrary coset $aH$ is the set $\{g\in G : gaH = aH\}$. We find that this is exactly the conjugate subgroup $aHa^{-1}$. More generally, if $G$ acts transitively on a set $S$ and $gs = s'$, then $\Stab(s') = g\Stab(s)g^{-1}\subseteq G$.

@@

\parenproclaim Theorem S (First Sylow Theorem).
